Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-06-22 doc/paper.Rmd
through results (survival, lag)

---
title: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
shorttitle: "Delayed effects of climate on vital rates"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - María Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false

@@

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This increased sensitivity to climate fluctuations, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

Whether the demography of plant populations in tropical forest fragments is more susceptible to drought and other climatic extremes remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [*e.g.,* @gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

@@

In 1997--1998 we established a series of 5000 m^2^ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest.
This study uses data only from 1-ha fragments and continuous forest sites because these represent extremes and because there are only three 10 ha fragment plots.

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
In this study, we used the product of shoot number and plant height as our measure of plant size.
Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
Plants that were not found for three consecutive surveys were considered dead.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
For additional details on the location of plots, survey methods, and *H. acuminata* population structure see @brunaHabitatFragmentationDemographic2002.

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
SPEI calculations were made using the `SPEI` package[@begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought sometimes differed (Figure S2; Table S1 <!--# update crossrefs manually -->).

@@

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

The `dlnm` package does not currently allow modeling of factor by smooth interactions.
This means we could not include habitat as an interaction term.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
Significant main effects of habitat type were assessed by looking for overlap in the 84% confidence intervals of model intercepts (84% CIs of two samples drawn from the same population overlap about 95% of the time [@paytonOverlappingConfidenceIntervals2003]).

@@

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d). The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the southwestern Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

@@

Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).

@@

*Survival:* The proportion of surviving plants across all plots was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d); it was also preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv}=0.95$; Figure \@ref(fig:eda)b), with 85% CI of model intercepts overlapping, indicating no significant difference between the two habitat types.
However, survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.

@@

*Growth:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were on average substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively), however, size was did not differ significantly between habitats (overlap in 84% CI of model intercepts).
The distribution of sizes of individuals was somewhat bimodal in both habitats and the major difference in plant size was in the larger of the two modes being smaller in fragments (Figure \@ref(fig:size-plot)d).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).

@@

*Flowering:* While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), although this difference did not reach statistical significance (84% CIs of model intercepts overlapped).
This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Even though flowering probability of large plants was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive size threshold in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

For 1-ha fragments, SPEI in the preceding 13 months had a significant effect on survival, with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
Wet conditions (SPEI \> 1) in the preceeding 23 months had a significant negative effect on survival in 1 ha fragments (Figure \@ref(fig:surv-spei)b).
In contrast, the effect of recent SPEI in continuous forest was weaker and significant for a shorter window of only the preceding 6 months (Figure \@ref(fig:surv-spei)a).
The effect of SPEI in continuous forest also differed in that it was unidirectional---survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)a).
Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census) and 32--36 months.
Drought (i.e., SPEI \< -1) 15--20 months prior to a census was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival (put another way, there was a nearly linear relationship between SPEI and survival probability).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first year of census data (1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.

While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, when the current wet season is unusually wet (SPEI\>2), plants in continuous forests were larger by up to log(size) = 0.57 due to reduced growth in fragments.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, but the magnitude of the response was greater in 1-ha fragments.

Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20 ) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In Fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).

>>>>

---
title: "Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments"
shorttitle: "Delayed effects of climate on vital rates"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - María Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false

@@

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with effects that are more pronounced on fragmented populations [@oliver2015; @holyoak2016]. While the demographic consequences are expected to be similarly severe in the tropics [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for species found in tropical biomes [@paniw2021]. Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015]. Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019]. This increased sensitivity to climate fluctuations, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

Whether the demography of plant populations in tropical forest fragments is more susceptible to climatic extremes remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions. Second, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those of plants in forest fragments [@lauranceEffectsStrongDrought2001]. Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011]. These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016]. They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [*e.g.,* @gagnonGrowthUnderstoryHerb2011]. Indeed, lagged responses of demographic vital rates to climate may be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

@@

This study uses data collected in four 1-ha fragment reserves and six continuous forest sites.
In 1997--1998 we established a 5000 m^2^ plots ($50 \times 100$m) in each of these sites in which we marked and measured all *Heliconia acuminata*; plots in 1-ha fragments were on one randomly selected half of the fragment, while plots in continuous forest were located 500--4000 m from the borders of secondary and mature forest.
The distance between plots ranged from 500 m--41 km.
Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).

Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
In this study, we used the product of shoot number and plant height as our measure of plant size.
Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
Plants that were not found for three consecutive surveys were considered dead.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
For additional details on the location of plots, survey methods, and *H. acuminata* population structure see @brunaHabitatFragmentationDemographic2002.

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993]. SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
SPEI calculations were made using the `SPEI` package [@begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought sometimes differed (Figure S2; Table S1 <!--# update crossrefs manually -->).

@@

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs; @gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

The `dlnm` package does not currently allow the modeling of interaction terms, which means we could not asses the interaction of habitat type and lagged effects.
We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
Significant main effects of habitat type were assessed by looking for overlap in the 84% confidence intervals of model intercepts; the 84% CIs of two samples drawn from the same population overlap about 95% of the time [@paytonOverlappingConfidenceIntervals2003].

@@

The meteorological droughts in our focal region indicated by SPEI are generally consistent with those reported in the literature (Table S1).
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d). The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the southwestern Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

@@

@@

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was apreceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.

@@

*Growth:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average `r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were `r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 34% larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).

Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).

@@

*Flowering:* The overall proportion of plants that flowered was very low.
While it was almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of large plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 13 months.
The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
In contrast, the effect of recent SPEI in continuous forest was weaker, with only the 6 months preceding a census having a significant effect on survival (Figure \@ref(fig:surv-spei)a).
These short-term effects of SPEI on survival were also unidirectional---the probability of survival declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
In contrast, the most pronounced negative effects of SPEI on the survival of plants in continuous forest were at lag times of 15--20 months and 32--36 months.
Drought 15--20 months prior to a census (i.e. two dry seasons prior to a census) was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival.
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first year of census data (1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.

*Growth:* While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, when the current wet season is unusually wet (SPEI\>2), plants in continuous forests were larger by up to log(size) = 0.57 due to reduced growth in fragments.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, but the magnitude of the response was greater in 1-ha fragments.

*Flowering:* Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20 ) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In Fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
